The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-09-30 paper.tex
Added SonAngle numbers to SonAngle hierarchy diagram

Fleischhacker's analysis rested on empirical determinations of sonority contour faithfulness, and did not attempt to determine such a relation. \citet{steriade.2006} proposed that input and output sonority contours should match in terms of whether they are rising or falling, and to what degree, but did not suggest a concrete mathematical relation.  \citet{flemming.2008} formalises Steriade's approach with the metric {\sc Sonority Rise}, the ratio of the gradients of the two contours.

@@

In this paper, I suggest that any such metric should bear two properties. The first is that suggested by previous analyses, that the more positive the gradient of the sonority contour from C\textsubscript{1} to C\textsubscript{2}, the smaller the faithfulness cost and hence the more likely epenthesis is to occur.
The second is that for a given sonority distance between C\textsubscript{1} and C\textsubscript{2}, the more sonorous C\textsubscript{1} and C\textsubscript{2} are, the more likely epenthesis is to occur.

@@

Syllable Contact forms the basis for the main existing analysis of Chaha epenthesis by \citet{rose.2000}.  I show that the faithfulness-based analysis, powered by the metric of {\sc Sonority Angle}, is slightly more economical in addressing the data. Furthermore, in the case of Irish, Syllable Contact makes incorrect predictions regarding the data.

@@

The observation with which we started was that the more steeply rising the sonority profile of a consonant cluster, the more likely the cluster to undergo epenthesis. Thus the absolute difference in sonority between C\textsubscript{1} and C\textsubscript{2} must be factored into the metric. To this, I add the claim that the more sonorous C\textsubscript{1}, the more likely the cluster is to undergo epenthesis.

These two factors are neatly captured by the metric {\sc Sonority Angle}, which is defined as the angle between the underlying C\textsubscript{1}C\textsubscript{2} sonority contour and the surface C\textsubscript{1}V contour:

@@

\vspace{-5em}
\begin{center}
\begin{tikzpicture} [shorten >=1pt,scale=0.33]
                      \draw [<-] (-3,12) -- (-3, 0.5) ; % axis
                      \node at (-3, 13.0) {Sonority}; % axis label
					 % left diagram
    \coordinate (A) at (0,4); % pivot - C1 - F * 3
    \coordinate (B) at (15,2); % C2 - T * 3
    \coordinate (C) at (15,12); % V
    \draw (B) -- (A) -- (C);
	\tkzMarkAngle[size=2.5cm](B,A,C);
	\tkzLabelAngle[pos=4](B,A,C){$\theta_1$};
    \coordinate (D) at (15,6); % another C2 - N * 3
    \coordinate (E) at (0,8); % another C1 - R * 3
    \draw (D) -- (E) -- (C);
	\tkzMarkAngle[size=2.5cm](D,E,C);
	\tkzLabelAngle[pos=4](D,E,C){$\theta_2$};

@@

We formalise the idea of the faithfulness cost by defining a family of \textsc{Ident} constraints that penalise outputs that incur faithfulness costs of greater than a certain $n$.

@@

This family of faithfulness constraints forms a stringency hierarchy. Any cluster that would, upon epenthesis, violate \textsc{Ident(SonAngle)}$<$n would also violate {\sc Ident(SonAngle)}$<$m where m$>$n.
The resulting hierarchy of clusters, ranked according to their resistance to epenthesis as defined
by their {\sc Sonority Angle}, is as follows.

@@

Notice that out of the falling sonority clusters, those that decrease in sonority by a single step
-- namely RN, NF and FT -- have smaller {\sc Sonority Angles} than the ones that have a greater fall
in sonority. Furthermore, out of these three clusters, the one with the most sonorous C\textsubscript{1}, RN, has the smallest {\sc Sonority Angle}. We thus predict that out of the falling sonority clusters, RN is the most likely to be broken up by epenthesis. The case study on Chaha will demonstrate that this is the case.

>>>>

Fleischhacker's analysis rested on empirical determinations of sonority contour faithfulness of specific clusters such as stop-liquid (TR) clusters and fricative-stop (FT) clusters, and did not attempt to determine a formula for arbitrary values of C\textsubscript{1}, C\textsubscript{2}. \citet{steriade.2006} proposed that input and output sonority contours should match in terms of whether they are rising or falling, and to what degree, but did not suggest a concrete mathematical relation.  \citet{flemming.2008} formalises Steriade's approach with the metric {\sc Sonority Rise}, the ratio of the gradients of the two contours.

@@

In this paper, I suggest that any such metric should bear two properties. The first is suggested by previous analyses, while the second is original to this paper.

\ex. Desiderata for a sonority contour similarity metric \label{desiderata}
     \a. The more positive the gradient of the sonority contour from C\textsubscript{1} to C\textsubscript{2}, the smaller the faithfulness cost.
     \b. For a given sonority distance between C\textsubscript{1} and C\textsubscript{2}, the more sonorous C\textsubscript{1} and C\textsubscript{2} are, the smaller the faithfulness cost.

@@

Syllable Contact forms the basis for the main existing analysis of Chaha epenthesis by \citet{rose.2000}.  I show that the faithfulness-based analysis, powered by the metric of {\sc Sonority Angle}, is equally capable of explaining the Chaha facts. Furthermore, in the case of Irish, Syllable Contact makes incorrect predictions regarding the data.

@@

The first desideratum with which we started was that the more steeply rising the sonority profile of a consonant cluster, the more likely the cluster to undergo epenthesis. Thus the absolute difference in sonority between C\textsubscript{1} and C\textsubscript{2} must be factored into the metric. To this, I add the claim that the more sonorous C\textsubscript{1}, the more likely the cluster is to undergo epenthesis.

These two criteria are fulfilled by the metric {\sc Sonority Angle}, which is defined as the angle between the underlying C\textsubscript{1}C\textsubscript{2} sonority contour and the surface C\textsubscript{1}V contour:

@@

\vspace{-9em}
\begin{center}
\begin{tikzpicture} [shorten >=1pt,scale=0.33]
                      \draw [<-] (-3,12) -- (-3, 0.5) ; % axis
                      \node at (-3, 13.0) {Sonority}; % axis label
					 % left diagram
    \coordinate (A) at (0,4); % pivot - C1 - F * 3
    \coordinate (B) at (15,2); % C2 - T * 3
    \coordinate (C) at (15,12); % V
    \draw (B) -- (A) -- (C);
	\tkzMarkAngle[size=2.5cm](B,A,C);
	\tkzLabelAngle[pos=4](B,A,C){$\theta_1$};
    \coordinate (D) at (15,6); % another C2 - N * 3
    \coordinate (E) at (0,8); % another C1 - R * 3
    \draw (D) -- (E) -- (C);
	\tkzMarkAngle[size=2.5cm](D,E,C);
	\tkzLabelAngle[pos=4](D,E,C){$\theta_2$};

@@

I formalise the idea of faithfulness cost by defining a family of \textsc{Ident} constraints that penalise outputs that incur faithfulness costs of greater than a certain $n$, following \citep{flemming.2008}.

@@

This family of faithfulness constraints forms a stringency hierarchy. Any cluster that would violate \textsc{Ident(SonAngle)}$<$n, would also violate {\sc Ident(SonAngle)}$<$m, where m$>$n.

The resulting hierarchy of clusters, ranked according to their resistance to epenthesis as defined
by their {\sc Sonority Angle}, is as follows.

@@

Notice that out of the falling sonority clusters, those that decrease in sonority by a single step
--- namely RN, NF and FT --- have smaller {\sc Sonority Angles} than the ones that have a greater fall
in sonority. Furthermore, out of these three clusters, the one with the most sonorous C\textsubscript{1}, RN, has the smallest {\sc Sonority Angle}. We thus predict that out of the falling sonority clusters, RN is the most likely to be broken up by epenthesis. The case study on Chaha will demonstrate that this is the case.
